---
title: "KP R Intro, Part 1: Intro to RStudio, Rmd files, data types, subsetting and importing/exporting"
author: "Nerissa Nance"
date: "March 12, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 
 

#Part 1 learning objectives

- Utilizing files in RStudio, creating an Rmd file, download, install packages and read them into your library, and read key documentation
- Reviewing different data types, naming/calling syntax, subsetting syntax and data structures
- Subsetting data
- Loading a csv file and SAS file into the environment 
 
(Relevant chapters in [R for Data Science](https://r4ds.had.co.nz/) : Chapter 1 Intro, Chapter 27 R Markdown, Chapter 20 Vectors)

#Introduction

This will be a review of the 'Part 0' training slides that you reviewed before this class.

### The basics: Navigating RStudio and Projects
Currently, you have the "KP-R-Intro-Workshop" project open (as you can see in the upper right-hand corner of your screen). 

Projects are handy because a project's folder is automatically set as the working directory, which allows us to not worry about explicitly setting the working directory. This can be more convenient when working on the same code across different computers. 

We can see recent projects by selecting File -> Recent Projects in the upper left corner of RStudio, or by clicking the list of recent projects in the upper right corner. If we click the name of the project it will switch to that project, or if we click the button it will open in a new session.

Let's try creating a new project.

* Create a new project: File -> New Project
* Choose new directory or existing directory as preferred.
* Click "open in new session" to have multiple RStudio projects open (if you're using PC RStudio).

Easy Challenge: close your project and open "KP-R-Intro-Workshop" back again.


### RStudio interface

When we start RStudio we have no files open by default. Click File -> New File -> RScript to start a new R script (or program). This is a text file that you can fill with code or comments(to comment, start the line with a #).

As an alternative to an R script we can use a **markdown file**. R Markdown files combine R scripts with simple text formatting, which makes reports  (PDFs or HTMLs) easy to generate with embedded R output. 

RStudio items to note:

* Files show up as tabs in the editor window, usually the upper left.
* The console is the R session where our commands are executed.
* Environment tab in the upper right, and history tab.
* Files tab in bottom right, plus plots, help, etc.
* We can click the "run" button to run the current R line in the editor.

### RStudio Script challenge

Find out your working directory by executing the following command:
```{r}
getwd()
```
 
### A note about Github and Gitlab

Github is a version control software that allows you to track how your code changes over time. Gitlab has many of the same features as Github. Kaiser has its own Gitlab enterprise account, which allows us 

You can get your Gitlab account by emailing Jingrong Yang. You can connect your Gitlab account to your RStudio by following instructions in this helpful happygitwithr [tutorial](https://happygitwithr.com/) 


### Basic R markdown file.

File -> New -> R Markdown

To write R code we need to start an R block with three backticks followed by r in braces (squiggly brackets - { and }). We can write commands inside of those brackets to change settings.

You can save that PDF version [here](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf).


### Package installation
Let's install some packages that we will use during the training. You can install packages by clicking the "Packages" tab in the bottom-right window, clicking install, and then searching for the package you wish to install.

Packages can also be installed through the console using `install.packages()`. The package name must be wrapped in quotation marks so that R knows it is searching for that particular package named `"swirl"`, rather than previously defined data named swirl.

> Packages only need to be installed once, though it is generally good to keep RStudio and your packages up to date. 

To install the packages we will use in the workshop: 
```{r, eval=FALSE}
install.packages(c("knitr", "rmarkdown", "swirl", "mlbench", "tidyverse", "rio"))
```

### The `library` function  
Before using a previously installed package, you must retreive it with `library()` when you begin a new R session. You do not need to reinstall packages each time you quit and restart your R session. 

```{r}
library(tidyverse)
library(knitr)
library(rmarkdown)
library(rio)
library(mlbench)
# these packages can now be used in our RStudio session!
```

### Getting help  
The `?` symbol can be used to bring up the help pages:  
```{r, eval=FALSE}
?update.packages
?read.csv
?summary
?describe
?plot

```


###Global environment

You define objects through `variable assignment` and they are stored in your `global environment` on the top right of the RStudio interface.

Your global environment is where all the objects (data sets, scalars, vectors, functions, etc.) are stored. It is 'global' because anything in it can be used throughout your session. 

### Object assignment (`<-`)
 Objects are stored in R's memory and can be retrieved ("called") when you need them.  

You define objects through variable assignment using the **assignment operator** `<-`. This is a "less than" `<` symbol immediately followed by a hyphen `-`

> You will also see that `=` is frequently used in place of the assignment operator. In most cases, they can be used interchangeably, but we recommend using the assignment operator for consistency.  

**Object definition/variable assignment requires three pieces of information**:  
1) object_name  
2) `<-`  
3) definition/assignment  

Let's define an object named `numeric_object` and define it as the number 4:
```{r}
numeric_object <- 4

# the ls() function will show us what is in our global environment
ls()


r <- 2

#Try an equation:

area <- pi * r^2

area
```

### Interacting with the global environment
We now have an object defined in our global environment!

"Call" (retrieve) the data contained wihtin the object by typing its name into your script and running the line of code
```{r}
numeric_object 
print(numeric_object)
# 4 is returned
```

Let's try another example, this time using character data. Note that character data is **always** contained within quotation marks `" "` 
```{r}
welcome_object <- "Welcome"
ls()
```

We now have four objects in our global environment. Call the data contained within the object by typing its name and running the line of code:
```{r}
welcome_object
```

### Naming R objects  
Object names can be anything, but are always **case sensitive**. However, they cannot begin with a number and generally do not begin with symbols. However you choose to name your objects, be consistent and use brief descriptions of their contents.  

Convention: we **highly** recommend that you do not use caps when naming objects. It makes it more cubersome to call objects later, and can lead to confusion and misnaming. You can also use either a '.' or '_' to seprate words in a name. In these examples, we will use '_', because this is more similar to SAS naming conventions. 

Names must be __**unique**__. If you reuse an old name, the old definition will be overwritten.  

Some words are already reserved by R (for more information, type `?Reserved` in your console), and some have stored default values (like pi) which can be overwritten if you choose (for more informaiton, see `?Constants`).

Let's overwrite the object `welcome_object` from above.
```{r}
welcome_object <- "welcome to class" 
welcome_object


welcome_object <- "welcome to part 1"
ls() 
```
See how the definition of `welcome_object` changed in your global environment window? However, there is still only one 'welcome_object' in your global environment. 


### Variable classes (`class`)
Each variable in R has a `class` that summarizes the type of the data stored within the object. We will talk more about data types below.

> NOTE! "numeric" is the default data type for numbers in R:

```{r}
# numeric (NOT integer) data type!
class(numeric_object) 
# character type
class(welcome_object) 

```

### Removing variables (`rm`)
We remove individual variables from our environment with `rm()`. For example, to remove `numeric_object`, we type:
```{r}
rm(numeric_object)
ls()
```
See how `numeric_object` disappeared from our global environment?

We can also wipe the entire environment with `rm(list = ls())` (or, click the broom icon in the upper right "global environment" pane)
```{r}
rm(list = ls()) 
ls()
```
Now, all objects are gone from our global environment.

### Vignettes
Double question marks ?? will lead you to coding walkthroughs called "vignettes". These usually come with preloaded data and examples:  
```{r}
??psych
```
You will often find that your questions are not answered by the help pages nor vignettes. In that case you should Google-search your question or error message along with the name of a free help website.  

For example, to get help making boxplots, I might Google search "R how to make boxplots stack overflow".




# Atomic data types in R
Numeric, character, and logical (aka "boolean") data types all exist at the `atomic level`. Normally this means that they cannot be broken down any further and are the raw inputs for functions in R. Other R variables are frequently built upon these atomic types.

### Numeric type
Numeric data are numbers and integers. You may also hear numeric data referred to as `float` or `double` data types. By default, R stores everything as `doubles` (64 bit floating point numbers) which makes R very memory hungry. 

Define an object called `num` and then check its class:
```{r}
num <- 2 * pi
num
class(num)
```

Standard mathematical operators apply to the creation of numeric data:
`+`   `-`   `*`   `^`   `**`  `/`  `%*% (matrix multiplication)`  `%/% (integer division)`  `%% (modular division)`
```{r}
5 + 2  
5 - 2  
5 * 2  
5 ^ 2  
# same as ^ :
5 ** 2  
5 / 2  
5 %/% 2 
5 %% 2  
5 %*% 2  
```

### Character type
Character (aka string or text) data are always contained within quotation marks `" "`. Character handling in R is fairly close to character handling in a Unix terminal.  

Let's create an object called `char` and define it with some character data. Then, check its class:
```{r}
char <- "This is character data"
char
class(char)
```



### Logical type  
Logical (boolean) data are dichotomous TRUE/FALSE values. R internally stores `FALSE` as 0 and `TRUE` as 1. Define a logical object:
```{r}
bool_object <- TRUE
bool_object
class(bool_object)
```
We recommend to always spell out `TRUE` and `FALSE` instead of abbreviating them `T` and `F` which might be mistaken for previously defined variables - therefore, you should not save anything as `T`, `F`, or `TRUE` or `FALSE`! Thus, there are certain words that do not make good variable names. R has many reserved words that you should avoid using. See `?reserved` for more information.

Note that logical data also take on numeric properties. Remember that `TRUE` is stored as the numeral 1 and `FALSE` is stored as 0. 
```{r}
bool_object + 1
TRUE - TRUE
TRUE + FALSE
FALSE - TRUE
```

### Logical tests  
Logical tests are helpful in R if you want to check for equivalence. Logical tests compare two objects and return a logical output. This is useful for removing missing data and subsetting (you will learn more about this in Part 2). Note the use of the double equals `==` sign. 

```{r, eval=FALSE}
?"=="
?"&"
?"|"
?"!"
```
```{r}
TRUE == TRUE # is equivalent
FALSE == FALSE
TRUE == FALSE
TRUE & TRUE   # and 
TRUE | FALSE  # or
TRUE != FALSE  # not
TRUE > FALSE # greater than
FALSE >= TRUE # greater than or equal to
```
> See the below tables to help you remember your logical operations and functions.  

[Brunsdon and Comber 2015, page 15](https://us.sagepub.com/en-us/nam/an-introduction-to-r-for-spatial-analysis-and-mapping/book241031) offer a useful summary table of logical operators in R. These are useful for comparing two data variables to each other:    

Logical operator | Description
---------------- | -----------
==               | Equal
!=               | Not equal
>                | Greater than
<                | Less than
>=               | Greater than or equal
<=               | Less than or equal
!                | Not (goes in front of other expressions) 
&                | And (combines expressions)
`|`              | Or (combines expressions)  

[Brunsdon and Comber 2015, page 102](https://us.sagepub.com/en-us/nam/an-introduction-to-r-for-spatial-analysis-and-mapping/book241031) also offer an excellent summary table of logical functions in R.


# **Challenge 1**
1. What is the three-piece recipe for variable definition? What piece goes where?  
2. Define two numeric variables.  
3. Define two character variables.  
4. How do you check what types of data these variables are?  
5. What does the `ls()` function do? Where else do you see this information?  
6. Remove one of your numeric variables.  
7. Try to subract your remaining character variable from your numeric one. What happens? What might this tell you about data of different types?  
8. Define a numeric object as 0 and check its class.  
9. Define a boolean object as "FALSE" and check its class.  
10. Use `==` to compare your numeric and boolean object. Are they equivalent? Why or why not?  
11. Wipe your environment clean.  
 
 
```{r}
## YOUR CODE HERE


```



### Data testing and type coercion (`as.type`)
Similar to `class`, you can check whether your data is a certain type using the `is.` commands.

"Type coercion" refers to changing the data from one type to another. You can change data types with `as.` and then the data type you wish to convert to.  
[Brunsdon and Comber 2015, page 102](https://us.sagepub.com/en-us/nam/an-introduction-to-r-for-spatial-analysis-and-mapping/book241031) also offer an excellent summary table of data types in R.

Some of the more common data types include:

Type        | Test          | Conversion
----------- | ------------- | ----------
character   | is.character  | as.character
integer     | is.integer    | as.integer
list        | is.list       | as.list
logical     | is.logical    | as.logical
numeric     | is.numeric    | as.numeric
double      | is.double     | as.double



### Type coercion example (`as.numeric`)  
```{r}
# Create some character data
char_data <- "9"
class(char_data)

# Coerce this character data to numeric data type
as.numeric(char_data)
class(char_data)
# What happened here? Why did `char_data` not change classes? (hint: we did not overwrite the object!)

char_data2 <- as.numeric(char_data)
class(char_data2)
char_data2
```

### Type coercion example 2 (`as.integer`)
You can change numeric type to integer type using `as.integer` 
```{r}
num <- 4
class(num)

int <- as.integer(num)
class(int)
int
```

Now, create some character data and try to convert it to integer type:
```{r}
# Create a new object
chartonum <- "three"
chartonum   #Note that the word three is contained within " " 
class(chartonum)

# What happens if we try to coerce character to numeric data type? 
as.integer(chartonum)
```
Why did this fail? Can R change character data to numbers? Why not? (hint: R has no protocol for automatically coerce words to numbers). As you can see, trying to coerce data types can lead to weird behavior sometimes. 

# **Challenge 2**
1. Create a character object and check its type using `is.character` and `class`. What is the difference between these two methods?    
2. Create a numerical object that takes a value that contains decimals. Now change ("coerce") this object into an integer value using the command `as.integer`. What happens, and why?

```{r}
## YOUR CODE HERE


```

# Data structures
There are several kinds of data structures in R. Data structures are collections of data objects (e.g., numeric, character, and logical vectors, lists, and matrices) that work together. These four are the most common:  

1. vector  
2. list  
3. matrix  
4. dataframe  

### Vector
A **VECTOR** is an ordered group of the same kind of data. "Ordered" means that their position matters. Vectors are one-dimensional and homogenous, and are thus referred to by their type (e.g., character vector, numeric vector, logical vector).  

Create a numeric vector by combining/concatenating elements with `c()`  
```{r, eval=FALSE}
?c
```
```{r}
numeric_vector <- c(3, 5, 6, 5, 3)
numeric_vector
```
This numeric vector contains five elements.  

You can index a vector using square brackets (more on this in the subsetting section of Part 2). For example, to see what value lives in the third position of the object `numeric_vector`, you could type: 
```{r}
numeric_vector[3]
```

You can also add items to a vector using `c()` and a comma `,` (as long as it is the same data type)
```{r}
numeric_vector2 <- c(numeric_vector, 78)
numeric_vector2
```

It doesn't matter what the datatype is for a vector, as long as it is all the same
```{r}
logical_vector <- c(TRUE, TRUE, FALSE, FALSE, TRUE)
logical_vector

logical_vector2 <- c(logical_vector, c(FALSE, FALSE, FALSE))
logical_vector2
```

You can also add and multiply vectors, but they need to be the same length
```{r}
logical_vector * logical_vector
```


### Generating random vectors (`seq`)
You might need to create vectors that are sequences of numbers. You can do this via `seq`. Here we specify a vector from zero to the `length` of our object `logical_vector2` (eight). The argument `by = 2` tells R that we want only the even numbers!
```{r, eval=FALSE}
?length
logical_vector2
length(logical_vector2)

seq(from=0,to=length(logical_vector2),by=2)
```

### Sequence where `by = 1` (`:`)
R also gives you a shorthand operator for creating sequences in whole number increments of 1. This is the colon symbol `:`
```{r}
0:8

sequence_object <- 28:36
sequence_object 

0:length(logical_vector2)
```

### `set.seed`    
You can also sample random groups of numbers. Use `set.seed()` to ensure that we all always get the same random draws from the parent universe, even on different machines
```{r}
?set.seed
?runif
?rnorm
?sample
```
Set the seed, and then choose our values:
```{r}
set.seed(2346)
# 20 random samples from uniform distribution between the numbers 3 and 7
uniform <- runif(20, 3, 7) 
uniform

# 20 random samples from the normal distribution with a mean of 0 and standard deviation of 1
normal <- rnorm(20, 0, 1) 
normal

# 20 random samples from between the numbers 5 and 10. Note that `replace=TRUE` signifies that it is OK to reuse numbers already selected.
integer <- sample(5:10, 20, replace = TRUE) 
integer

character <- sample(c("Cat", "Dog", "Pig"), 20, replace = TRUE) # 20 random samples of character data
character


```

### List
A **LIST** is an ordered group of data that are not of the same type. Lists are heterogenous. Instead of using `c()` like in vector creation, use `list()` to create a list:
```{r, eval=FALSE}
?list

list1 <- list(TRUE, "one", 1) # include three kinds of data: logical, character, and integer
list1
class(list1)
```

Lists are simple containers and are not additive or multiplicative like vectors and matrices are:
```{r, eval=FALSE}
list1 * list(FALSE, "zero", 0) # Error
```

### Matrix  
**MATRICES** are homogenous like vectors. They are tables comprised of data all of the same type. Matrices are organized into rows and columns. 

Use `matrix()` to create a matrix
```{r, eval=FALSE}
?matrix
```
We can also specify how we want the matrix to be organized using the `nrow` and `ncol` arguments:
```{r}
matrix1 <- matrix(1:12, nrow = 4, ncol = 3) # this makes a 4 x 3 matrix
matrix1
class(matrix1)
```
We can also coerce a vector to a matrix, because a vector is comprised of homogenous data of the same kind, just like a matrix is:
```{r}
# Create a numeric vector from 1 to 20
vec1 <- c(1:20)
vec1
class(vec1)

# Coerce this vector to a matrix with 10 rows and 2 columns:
matrix2 <- matrix(vec1, ncol=2)
matrix2
class(matrix2)
```

### Data frame  
_**It is worth emphasizing the importance of data frames in R!**_ Inside R, a **DATA FRAME** is a list of equal-length vectors. These vectors can be of different types. Data frames are thus heterogenous like lists.  

This is simply a spreadsheet!  

Let's create a dataframe called `animals` using the vectors we already created:

We do this using `data.frame()`
```{r, eval=FALSE}
?data.frame
```
```{r}
animals <- data.frame(uniform, normal, integer, character, stringsAsFactors = FALSE)
# NOTE: `stringsAsFactors=FALSE` means that R will NOT try to interpret character data as factor type. More on this below. 
```

Take a peek at the `animals` data frame to see what it looks like:
```{r}
head(animals)
```

We can change the names of the columns by passing into it a vector with our desired names
```{r}
# Create a vector called `new_df_names` with the new column names and pass this vector into `colnames()`
colnames(animals) <- c("Weight", "progress", "height", "type")
head(animals)
```

We can check the structure of our data frame via `str()`
```{r, eval=FALSE}
?str
str(animals)
```
Here, we can see that a data frame is just a list of equal-length vectors! For readability purposes, `str()` displays columns from top to bottom, while the data are displayed left to right. 

##### Learning about your data frame  
```{r}
# View the dimensions (nrow x ncol) of the data frame:
dim(animals) 

# View column names:
colnames(animals)

# View row names (we did not change these, so they default to character type)
rownames(animals)
class(rownames(animals))
```


##### Factor data type  
Factor data are categorical types used to make comparisons between other data. Factors group the data by their "levels" (the different categorical groups within a factor).  

For example, in our `animals` dataframe, we can coerce "Type" from character to factor data type. Cat, Dog, and Pig are the factor levels. If we might want to compare heights and weights between Cat, Dog, and Pigs, we set the character "Type" vector to factor data type. We can do so with `factor()`.  

The dollar sign operator `$` is used to call a single vector from a data frame. 

```{r}
# "Name" is character data type. See how each column name is preceded by `$`?
str(animals)   
 class(animals$type)
animals$Type <- factor(animals$type)
# "Name" is now factor data type!
str(animals)  

```

Notice that R stores factors internally as integers and uses the character strings as labels. Also notice that by default R orders factors alphabetically and returns them when we call the "type" vector.
```{r}
animals$type
levels(animals$type)
```

##### Changing factor "levels"
Each animal type (Cat, Dog, and Pig) within the factor `animals$type` vector are the factor levels. 

If we want to change how R stores the factor levels, we can specify their levels using the `levels()` argument. For example:
```{r}
animals$type  

# What if we want to change the factor level sort to Levels: Dog Pig Cat?
animals$type <- factor(animals$type, levels = c("Dog", "Pig", "Cat"))

# Now when we call animals$Name, we can see that the levels have changed
animals$type 
```

 
 
#Data manipulation

In R, there are many ways to do the same thing. "Base r" is the built-in language of R, and it allows you to perform data manipulation and visualization. The syntax for base r can be a little confusing/daunting at times, however. In part because of this, Tidyverse was born. Tidyverse (developed by Hadley Wickham and colleagues) is a suite of packages that we loaded into our environment at the beginning of this Rmd file. These packages allow for easy data manipulation and visualization, and are often preferred by programmers. The Syntax is relatively intuitive, and consists of several commands that we will review during this training:

- filter
- select
- summarize
- mutate
- join

You can downlaod a handy cheatsheet dplyr, which is the primary package of tidyverse that handles data manipulation, here[https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf].

##Subsetting data with dplyr

Subsetting is often necessary when you want to perform a task on part of your dataset. You can subset using the 'filter' command in dplyr. We can also subset the data by selecting columns, using the select command. 

A note about chaining commands with the forward pipe operator '%>%':
part of the magrittr package of tidyverse, this command can make code easier to write and read. We'll show an example here.
```{r}

pigs_only <- animals %>% filter(type=="Pig")
head(pigs_only)

#without the pip operator, it would look like this:
pigs_only <- filter(animals, type=="Pig")


#Now let's subset to the type and height columns only:
typeheight <- animals %>% select(type, height)

#check out the columns names using colnames()
colnames(typeheight)


#we can also do both of those things in the same line:
animals %>% select(type, height)  %>% filter(type=="Pig")

```

>NOTE: as you see above, when you chain using the pipe operator, each object being piped is used as the FIRST argument of the next statement.


###Using logical tests to subset

You can any of the operators from the "logical tests" section above to subset your data. Here are a few more examples (note that the dataset is not being saved to an object, so it is only printed i nthe console):

```{r}

#Dogs OR Cats
animals  %>% filter(type=="Dog"|type=="Cat")

#Animals with a weight of at least four pounds
animals  %>% filter(height>=4)

#Dog with height over 7
animals %>% filter(type=="Dog" & height>7)

```


# **Challenge 3 **
1. Load the `iris` dataset! Type `data(iris)` to load it.  
2. What is this dataset? How do you find out?  
3. Subset the `iris` dataset:

    * to a data frame named "vv" with only the observations from the "versicolor" or "virginica" species 
    * to a data frame with only the Petal.Length and Petal.Width variables, where width is less than 3 and length is greater than 3.
    
Please do not copy and paste! Typing it out will help learn. :)
    
Extra credit #1: Try also subsetting to a dataset that only has length values greater than the median length OR width values greater than the median width (hint: you can use `summary()` to peek at descriptive stats)
Extra credit #2: Can you do the same thing but without hard coding the value?
   
```{r}
## YOUR CODE HERE


```


#Importing and Exporting



###Importing and exporting csv data files using `read.csv` and `write.csv`

Check first where your working directory is by typing:
```{r}
getwd()
```
(note: since we're using R projects, you should already know where your wd is, because it's set up automatically).


You can import data using the read.csv command. Take a moment to look at the help file for syntax tips:
```{r, eval=FALSE}
?read.csv
```
```{r}
df.who <- read.csv("./data/who_suicide_statistics.csv", 
                    header = TRUE, 
                    stringsAsFactors = FALSE,
                    na.strings = " ")
dim(df.who)
head(df.who)
str(df.who)


```

> Notice that `stringsAsFactors = FALSE`. If set to `TRUE`, R will try to guess which **character** vectors should automatically be converted to factors. This is problematic because 1) R is not always good at guessing and 2) R defaults to alphabetical and increasing numeric factor level sorting. This might not matter for your data, but it is recommended to set `stringsAsFactors = FALSE` and manually convert your desired character vectors to factors.

> `header = TRUE` will include the header row; `header = FALSE` will turn your header row into the first row of actual data. `na.strings = c(" ", 999)` indicates that blank cells and cells coded 999 allows you to specify data that should be automatically converted to `NA` upon importation. We do not have any here, so nothing is altered. 


###Importing and exporting with the Rio package

The Rio package simplifies importing/exporting by putting all data-specific import tools under one umbrella (popularly known as the "swiss army knife" of data importing and exporting)
 See [this helpful rio documentation as well](https://cran.r-project.org/web/packages/rio/vignettes/rio.html). 
 
```{r installing, eval=FALSE}
install.packages("rio")
?rio
```
```{r}
library(rio)

df.who2 <- import("./data/whostats_suicide.sas7bdat")
 dim(df.who2)
 head(df.who)
 
#Are these dataframes the same? Let's see:
 all.equal(df.who,df.who2)

 
 class(df.who$suicides_no)
class(df.who2$suicides_no)
```
>Moral of the story: beware formatting, esp. variable class changes!

# **Challenge 4**
1. Save the "animals" data frame as a .csv file named "animals.csv". 
2. Import the "animals" data frame using the rio package and the read.csv command. 
3. Compare the two data frames. Are they the same? Why or why not?
```{r}
## YOUR CODE HERE

```




# Subsetting in base R (`$`)

The dollar sign operator `$` will extract only a single vector/column within the data frame:
```{r, eval=FALSE}
?"$" # Remember to wrap symbols in quotation marks to view their help pages
```
```{r}
#Returns only the "age" vector from the `df.pid` data frame.
head(animals$type) 

```


#Visualize data (Part I)

We'll use this more in part 2, but take a moment to look at the data you imported. You can use the `hist()` command to do a histogram and the `summary()` command to look at summary statistics (using the `$` to subset to the variable you are interested in). For example:

```{r plotting}
names(df.who)
hist(df.who$suicides_no)

summary(df.who$suicides_no)

plot(df.who$year, df.who$suicides)

```

#**Challenge 5** (Extra credit)

* Create a NEW markdown file called "test"
* Add a header with the title "R Training Day 1" with size 1 (aka "header 1")
* In the first chunk of code, load the dataset mtcars (hint: `data(mtcars)`)
* Read about the dataset in the help file `?mtcars`
* In the second chunk of code, graph a scatterplot of weight and mpg--what does this tell you about the relationship between the two variables? Write dow na few sentences of explanation in the notes section (i.e. the white area)
* Compile as html.


# **Challenge 6** Homework (optional)
swirl is a package that helps you learn R by using R. Visit the [swirl](http://swirlstats.com/) homepage to learn more
```{r, eval=FALSE}
library(swirl)
swirl()
# follow the instructions until you can select number 1 "R Programming: The basics of programming in R"
```

> NOTE: type `bye()` to exist swirl. 

Acknowledgements
- [Wickham H. 2014. Advanced R](http://adv-r.had.co.nz/)  
- [Lander J. 2013. R for everyone: Advanced analytics and graphics](http://www.jaredlander.com/r-for-everyone/)  
- [Matloff N. 2011. The art of R programming: A tour of statistical software design](https://www.nostarch.com/artofr.htm)  

A lot of the source material for this training came from D-lab training materials authored by D-Lab staff. The original content can be accessed here: [https://github.com/dlab-berkeley/R-Fundamentals]
D-Lab works with Berkeley faculty, research staff, and students to advance data-intensive social science and humanities research. They're really cool, check them out!
 [D-Lab Calendar](http://dlab.berkeley.edu/calendar-node-field-date)  
 [D-lab Working Groups](http://dlab.berkeley.edu/working-groups)  
Some of this also came from Chris Kennedy's "Concise R" series, accessed [here](https://github.com/ck37/concise-r)
 
