---
title: 'KP R Intro, Part 2: More data manipulation, summarization and graphing'
author: "Nerissa Nance"
date: "September 5, 2018"
output:
  html_document: 
    number_sections: yes
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---



# **Challenge 1**
1. Clear your working environment
2. Call the `rio` package and load in the SAS file of the WHO data (look in the "data" folder for this project to see the name of the file; hint: `?rio` for notes on syntax) 

Extra credit #1: look at the data. How is it set up? why might this be challenging for analysis?
Extra credit #2: load in also the .csv file of the same dataset and compare the datasets using the `all.equal()` command. Are they the same?

```{r review, eval=FALSE}

##YOUR CODE HERE


```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("mlbench")
library(mlbench)
library(rio)
library(tidyverse)
```

Day 2 learning objectives:  

* 1. More dplyr: summarizing, merging  
* 2. Missing data (NA)  
* 3. Plotting with ggplot2 

 

# Loading data from files

Load the PimaIndiansDiabetes2 dataset from the mlbench package. This dataset is from the National Institute of Diabetes and Digestive and Kidney Diseases and contains health data for 768 patients.

```{r}

data(PimaIndiansDiabetes2) 
#I like to rename it so it's easier to type
assign("df_pid", PimaIndiansDiabetes2)
#can compare them using the "all.equal" command, just to double check
all.equal(df_pid, PimaIndiansDiabetes2)

# Read background information and variable descriptions.
?PimaIndiansDiabetes2

rm(PimaIndiansDiabetes2)

```


# Inspecting the data frame
Remember from Part 1 that we can learn a lot about data in R. For dataframes, the following functions are useful:
```{r, eval=FALSE}
# returns the structure of the dataframe
str(df_pid) 

# dataframe dimensions
dim(df_pid)    

#column names
colnames(df_pid)   
names(df_pid)    

#rownames
rownames(df_pid)  

# number of rows
nrow(df_pid)  
# number of columns
ncol(df_pid)  
# show rows with unique data
unique(df_pid)   

# if you want to know how many unique values there are, you can use this:
length(unique(df_pid$age))

# show the first "n" rows
head(df_pid, n = 10)
# show the last "n" rows
tail(df_pid, n = 4) 
```



# **Challenge 2**
1. Practice by extracting `pressure` and `glucose` from the `df_pid` dataframe using the dollar sign `$` and plot them in a histogram using the `hist()` function.

Extra credit #1: try this also using the generic `plot()` command.

```{r}
## YOUR CODE HERE



```


# Data summarization with base r 

We can view frequencies for of categorical data with `table`, and summary statistics of a continuous variable using `summary`. 
```{r}
# get frequencies for one variable:
table(df_pid$diabetes) 

#get a cross tabulation of two variables:
table(df_pid$diabetes, df_pid$pregnant) 

#get summary stats of a continuous variable:
summary(df_pid$glucose)
```



# Data summarization with dplyr

dplyr is a powerful tool for summarizing data. 

You can also use the `select()` command to look only at a specific column:

```{rsum dplyr }


df_pid %>% select (glucose) %>% summarize(mean_g=mean(glucose))

#need to tell dplyr to remove NAs!

df_pid %>% select (glucose) %>% summarize(mean_g=mean(glucose, na.rm=TRUE))

# let's look at the distribution of all the numeric variables at once using the `summarise_if` function of dplyr:
head(df_pid)
summarise_if(df_pid, is.numeric, funs(mean,median), na.rm = TRUE)

```

##Using `group_by()` to group your data with dplyr

Sometimes you want a certain statistic by group. To do this, you can  group by a certain variable, like diabetes status, using the 'group_by()' command
```{r}
 df_pid %>% group_by(diabetes) %>% summarise(mean_gluc=mean(glucose,na.rm = TRUE), mean_pressure=mean(pressure,na.rm = TRUE))

```
This gives you two statistics for each group: mean glucose and mean blood pressure among those with and without diabetes. What do you notice?

#Creating new variables with dplyr

The mutate() command in dplyr allows you to create new variables in your dataset. Let's say, for example, we wanted to create a ratio between blood pressure and glucose levels:

```{r}

df_pid %>% mutate(gbp_ratio= pressure/glucose)

#this prints the data, but we need to assign it in order to save it:
df_pid2  <- df_pid %>% mutate(gbp_ratio= pressure/glucose)


summary(df_pid2$gbp_ratio)

```



# Missing data (`NA`)
Identifying missing data can be important for subsetting purposes. R codes missing values as `NA`. Identifying missing data is important because dealing with it might be necessary to run basic tests like `mean()`. 
```{r, eval=FALSE}
?NA

# Scroll down to `na.rm`:
?mean 
```
```{r}
mean(df_pid$pressure) 
# This returns NA because R is unsure how to deal with NA cells for the `mean` computation.


table(is.na(df_pid$pressure))
```

Look at the "Usage" portion of the help file - we know what `rm` is, and we now know what `NA` is, so what do you think the logical argument `na.rm =` will do? That's right! We can use `na.rm = TRUE` to properly calculate the mean of the NonD column by now excluding the NAs. 
```{r}
mean(df_pid$pressure, na.rm = TRUE) 

#Now `mean()` returns the mean!
```

While `na.rm()` nor `str()` will not tell us which data are missing in a convenient way, `is.na()` does. Wrap the name of your data frame in `is.na()` to return logical values. Missing data is coded as `TRUE`, while present data are coded as `FALSE`
```{r, eval=FALSE}
?is.na
colSums(is.na(df_pid))

#or for a specific variable 
table(is.na(df_pid$pressure))


```


## Recoding missing data
Make a copy of `df_pid` named df.copy so we do not alter the original, then recode NA values and replace them with the median value:
```{r}
df_copy <- df_pid

df_copy$pressure[is.na(df_copy$pressure)] <- median(df_copy$pressure, na.rm=TRUE)
  
table(is.na(df_copy$pressure))


```


We can also subset only rows without any missing data using bracket notation. `complete.cases()` will find rows with no missing values.
```{r, eval=FALSE}
?complete.cases
```
```{r}

s_complete <- df_pid %>% filter(complete.cases(df_pid))

dim(s_complete)



#try this with one variable
s_complete2 <- df_pid %>% filter(complete.cases(pressure))  
dim(s_complete2)

```





#**Challenge 3**

1) Import "./data/whostats_suicide.sas7bdat" into your environment
2) Keep only the statistics for Albania for 2008
3) Summarize the WHO suicide numbers and population numbers by gender 
(Note: Look at the data! It's given by age group, so we will need to create a sum of these by using the `summarize()` function and the `sum` command)
4) Save the output as a .csv file

Extra credit #1: create a new variable for the proportion or percent of suicides for each group relative to the population.
Extra credit #2: summarize this new variable by gender.

```{r}
#YOUR CODE HERE

```


# Merging data with `cbind()` and `rbind()`
Merging data is useful when we want to combine two different dataframes that share a vector/column.   

`cbind()` will bind two data frames by their columns. Let's cbind `df1` and a new dataframe, `df3`:
```{r, eval=FALSE}
?merge
?cbind
?rbind 
```
```{r}


#let's say that we have this other column of data we need to add 
#(FYI: runif randomly chooses from a uniform distribution between the specified min and max. I'm making it the same size as the dataset for the sake of this example.) 
df3 <- round(runif(768, min=0, max=10), 2)
head(df3)

df_cbind <- cbind(df_pid, df3)
head(df_cbind)
dim(df_cbind)
```
> NOTE: If you are using cbind, because it binds the data by row names, you need to make sure your data is SORTED in EXACTLY the same way! 

To make sure your data is ordered the way you need it to be, you can order your data using the `order` command. For example, sort `df_pid` by ascending pressure values: 
```{r, eval=FALSE}

new_ordered <- df_pid[order(df_pid$pressure),] 
head(new_ordered)

#for descending values:
new_ordered2 <- df_pid[order(-df_pid$pressure),] 
head(new_ordered2)
```

`rbind()` will add more rows to an existing data frame. An example:
```{r}
 
#NOTE: datasets need to have the same number of columns
df.a <- df_pid[,c("diabetes", "age", "pressure")]
names(df.a)
df.b <- data.frame(diabetes = c("pos", "pos", "pos", "neg"),
                  age = c(44, 49, 50, 49),
                  pressure = c(60,70,60,75))
df_rbind <- rbind(df.a, df.b)
dim(df_rbind)
```
> NOTE: for rbind to work, you should have the same column names for both of your data frames! 

#Merging data using dplyr

You can merge data in R using a variety of tools, but one way is to use the package `dplyr`. dyplyr is a popular package for data manipulation. The commands for merging are very similar to what you use in sql. Take a moment to google `dplyr cheat sheet` to find a breakdown of the different commands.

Let's practice a join:

```{r}
library(dplyr)


#first make an indidivual identifier for the data, for the sake of example
df_pid$id <- rownames(df_pid)
#let's first split the data:
df1 <- df_pid[,c("id","pregnant", "glucose" , "pressure")]
df2 <- df_pid[1:400,c("diabetes", "id")]

joined <- dplyr::left_join(df1, df2, by="id")

dim(joined)
table(is.na(joined$diabetes))


```

#**Challenge 4**

1. Combine df1 and df2 again using dplyr, but this time using the inner_join command. What is different from when you used left_join, and why? How can you tell?

Extra credit #1: create 5 new rows of the df_pid dataset and append them to df_pid dataset using rbind(). (Be sure to create a NEW object so you are not overwriting your own!)

```{r}
#YOUR CODE HERE

```



# ggplot2
The "ggplot2" R package is a powerful way to graph your data. The syntax is additive, it uses a plus-symbol `+` to connect different arguments together.  

The [ggplot2](http://ggplot2.org/) R package eases the burden of plotting using base functions. Instead, it favors simplified syntax, which is different from anything you have seen thus far. ggplot2 is part of the Tidyverse.

You can use the analogy of a recipe, the `+` connects the different ingredients together to form your final product. Some "ingredients" are necessary, while others are supplementary.

The `ggplot` function is the main function call. You need three things main "ingredients" to create a ggplot:  
1) a dataset - some data to be visualized  
2) "aes"thetics - definitions of your coordinate system, point color and shape mappings, etc.  
3) geoms - how your data should be represented: points, bars, densities, ribbons, etc.

ggplot2 functions work in **layers**. Each time you wish to add a geom, title, axis label, or background color, you must add a new layer by typing the plus `+` sign.  

Check out the help file and the cheatsheet[https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf] and then see the below examples using the "iris" dataset:
```{r}

?ggplot2
data(iris)
```

# ggplot histogram 
Remember how we said the `ggplot` function works in layers? Look what happens if we specifiy our data set ("iris") and our global aesthetic ("aes"), but we do not specify our "geom". We get the coordinate system and a defined X axis (Petal.Length), but no representation of the data!
```{r}
ggplot(data = iris, aes(x = Petal.Length))
```

We can add the bars by adding a new layer with `+`, and then specify "geom_histogram()" to get produce the histogram! 

```{r}
ggplot(data = iris, aes(x = Petal.Length)) + 
    geom_histogram()

```

You can also change the color of the bars by adding the `fill` argument to the global aes, change the background color by adding `theme_` in a new layer, and the title by adding `ggtitle` in a new layer. The `bins` argument will change the number and width of the histogram bars.
```{r}
ggplot(data = iris, aes(x = Petal.Length)) + 
    geom_histogram(fill = "blue", color = "black", bins = 10) + 
    theme_classic() + 
    ggtitle("This is a ggplot histogram") + 
  ylab("Count")
```

# ggplot boxplots
Boxplots continue to follow this logic. However, this time we should specifically define our X axis as the factor variable we wish to investigate, and the Y axis as some numeric value: 
```{r}
box1 <- ggplot(data = iris, aes(x = Species, y = Petal.Length, fill = Species)) + 
    geom_boxplot() + 
    theme_minimal() + 
    ylab("Petal Length (cm)") +
    ggtitle("This are ggplot boxplots") 
#+guides(fill = FALSE)

box1
```
What happens if add in `guides(fill = FALSE)` in the above chunk? 

You can also add the "jittered" data points to get a better sense of the distribution (alpha is transparency):
```{r}
box1 + geom_jitter(color="grey", alpha=.5)

```
 

# ggplot scatterplots
Scatterplots work similarly. However, this time we must define both our X and Y axes in our global "aes" and select "geom_point()" as our geom. 

We can then automatically map the factor variable ("Species") to point types and shapes in the `color` and `shape` arguments. By doing so, a legend is automatically created.

The other layer arguments apply as well. `size` will change the size of the points, while `alpha` will change their transparency. 

We can also include a second `theme` argument layer where can can specify the legend position via `legend.position`. 

The X and Y axis labels can be changed via `xlab` and `ylab` layers, for example if we want to include our units of measurement (centimeters):
```{r}
petalplot1 <- ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + 
    geom_point(size = 10, alpha = 0.5) 
petalplot1

#adding some extra formatting:
petalplot2 <- petalplot1 + 
    theme_classic() +
    ggtitle("This is a ggplot scatterplot") +
    theme(legend.position = "top") + 
    xlab("Petal Length (cm)") + 
    ylab("Petal Width (cm)") + 
    theme(legend.position = "top",
          plot.title = element_text(hjust = 0.5, size = 40),
          legend.title = element_text(size = 20),
          legend.text = element_text(size = 20),
          axis.text.x = element_text(angle = 45, hjust = 1)) 

petalplot2

#adding a best fit line:
petalplot2 + geom_smooth(color="grey",method='lm',aes( group=1))
```

Setting group=1 in the new geom object overwrites the previous grouping for the line. Try removing it and see what happens!


# **Challenge 5** 
Try this on your own with the df_pid dataset!

1) Create a boxplot of pressure values by diabetes status
2) Do you get a `warnings` message? Why?
3) Add color, if you didn't yet in step #1
4) Add labels to the x and y axes 

Extra credit #1: fix the issue pointed out in the warnings() message (hint: use the code from the missing data section above)
Extra credit #2: try it on your own! Think about the data you have and what would be interesting to visualize. Refer to the ggplot2 `cheatsheet`, if you haven't already for inspiration.


```{r}
##YOUR CODE HERE


```

 
 
 
###Citations:
A lot of the source code for this material came from D-lab training materials authored by D-Lab staff. The original content can be accessed here: [https://github.com/dlab-berkeley/R-Fundamentals]
Contributions by Dlab staff: Evan Muzzall, Shinhye Choi, Rochelle Terman, Dillon Niederhut, Sam Abdel-Ghaffar.

